---
title: "EDA 5 - Updated habitat and swim bladders"
author: "Erin LaBrecque"
date: "11/20/2020"
output: html_document
---

This is file 5 of an exploratory data analysis (EDA) of the NEFSC Bottom Trawl Survey catch counts from 2008 through 2019.

EDA 1 cleans the species data using FishBase and SeaLifeBase.  
EDA 2 lists out all the species names, does some more name cleaning, and combines the species information with the trawl count data.
EDA 3 queries FishBase for swim bladder information (but I do not trust it) and SeaLifeBase and FishBase for habitat information.  
EDA 4 finds the spatial mismatches between strata listed in the survey data and the strata shapefile, and provides some basic count plots.

This EDA cleans up the habitat and swim bladder section. This EDA is a more "manual" than "programmatic" because I cannot find a database that can be accessed to retrieve this information. 


```{r setup, results='hide', message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      fig.width = 7, fig.height = 6.5)
library(knitr)
library(tidyverse)
library(lubridate)
library(kableExtra)
# library(sf)
# library(leaflet)
```


# Loading the data  
These are the outputs from the first and third EDA files. No data files from EDA 4 were manipulated or saved as updated files. For the observations that could not be classified to species, `non_sp_taxa_edit_sb_hab.csv` (should be in the `ouput_csv` folder) is a manual update of the swim bladder data. 
```{r readin}
dta <- read_rds("output_input/dta.rds") # original BTS trawl data (I always keep as a reference)
dta_ed <- read_rds("output_input/dta_ed.rds") # original BTS trawl data in tidy format, 0 counts removed

sp_fish <- read_rds("output_input/sp_fish_2.rds") # fish species information
sp_sealife <- read_rds("output_input/sp_sealife_2.rds") # all non-fish species information

non_sp <- read_rds("output_input/non_sp.rds") # fish and not-fish that could not be classified to species
#non_sp_habitat <- read_rds("output_input/non_sp_habitat.rds") # wide form of fish and non-fish that could not be classified to species with counts of species (re: habitat) for each taxa classification 

#manual_non_sp <- read_csv("output_csv/non_sp_taxa_edit_sb_hab.csv") # manual addition of swim bladder information of non-species observations
```

## Habitat frequency distribution of BTS observations not classified to species.  

There is probably an easier way to do this, but I am writing all this out step by step to keep things clear in my mind.   

Any observation in the BTS data that was classified to species now has a habitat designation (see the field `DemersPelag` in `sp_fish` and `sp_sealife` or look back at EDA 3). In the following section, I separate out the taxa for organisms that were not identified to species. This list is then split into two groups. 1) Those taxa that have a species representative, and 2) those taxa that do not have a species representative. 

An example of a species representative: the Family Apogonidae (cardinalfishes) is listed as an observation in the BTS data. The twospot cardinal fish, *Apogon pseudomaculatus*, is in the Family Apogonidae and also listed as an observation in the BTS data. This species is what I would call a "representative species" for the Family observation. In this case, it is the only species in the data with the Family Apogonidae, but there are other instances of Genera, Families, Classes, and Orders that have more than one species representative in the data. *Apogon pseudomaculatus* and Apogonidae were all collected in the same strata and during the same months, so I am going to assign the habitat for *Apogon pseudomaculatus* to the observation of Family Apogonidae. For instances where there is more than one species representative for a taxon, I will look at the strata, months, habitat, and breakdown of catch counts of each species to see if there are habitat weights that can be assigned and will break the trawl count data according to those weights.   

I am not quite sure what to do for taxa that do not have a species representative, yet.

Because there are multiple taxa levels involved, I wrote a two functions to look at what was and was not a species representative by taxa level. It was easier to break them out by the "having species representative" and "not having species representatives". 

*Side note:*
There is a great, little custom operator (`%nin%`) in the code block below for "not in". When using look-up lists and tables, a lot of filtering or subsetting is done by comparing lists - are values in one list in another list? To do that in R, you use: `%in%`. It is a logical operator (output is `TRUE` or `FALSE`), that provides the means to filter or subset observations. You can negate `%in%` (flip the filter) by adding `!` to the variable preceding `%in%`, but sometimes you need the answer the question, "what is not in the list?", and that can be different than negating. There is nothing hardwired into R for not-in, that's where the custom operator, "not in", comes in.

```{r function_get_spkids}
non_sp %>% 
  count(level) %>% 
  kable() %>% 
  kable_styling()

# need to add the "sp" back into BTS_sp and keep the misspellings to filter the 
# BTS count data 
select_list <- non_sp %>% 
  select(BTS_sp, Com_name, DB_sp, level, type) %>% 
  mutate(BTS_misspl = case_when(level == "Genus" ~ paste0(BTS_sp, " sp"), 
                            TRUE ~ BTS_sp))

# combining fish and non-fish data that has all the habitat information
# also combing the habitat options into pelagic, benthic, demersal, and reef-associated
both_fish_sealife <- bind_rows(sp_fish, sp_sealife) %>% 
  mutate(habitat = case_when(DemersPelag %in% c("pelagic", "bathypelagic", 
                                                "pelagic-neritic", "pelagic-oceanic") ~ "pelagic",
                             DemersPelag %in% c("benthic", "benthopelagic") ~ "benthic",
                             DemersPelag %in% c("demersal", "bathydemersal") ~ "demersal",
                             TRUE ~ DemersPelag))

# quick function for "not in"
`%nin%` = Negate(`%in%`)

# function to find species in the BTS data associated with a higher taxon level of 
# observation that could not be ID'd to species. 

get_sp_in_taxon <- function(taxon_filter, col_nm, ls_index){
  col_nm <- enquo(col_nm)
  ls_index <- enquo(ls_index)
  
  # this is for the join, roder here matters
  by = set_names(quo_name(ls_index), quo_name(col_nm))
  
  ls <- select_list %>% 
    filter(level == taxon_filter)
  
  ls_in_BTS <- both_fish_sealife %>% 
    filter(!!col_nm %in% ls$DB_sp) %>% 
    mutate(level = taxon_filter,
           inBTSdata = "in") %>% 
    left_join(select(ls, !!ls_index, BTS_misspl), by = by)
}


# another function to find the upper level taxa that do not have species 
# representation in the BTS data
get_sp_notin_taxon <- function(taxon_filter, col_nm){
  col_nm <- enquo(col_nm)
  
  ls <- select_list %>% 
    filter(level == taxon_filter)

  ls_in_BTS <- both_fish_sealife %>%
    filter(!!col_nm %in% ls$DB_sp) %>%
    select(!!col_nm) %>%
    distinct() %>%
    unlist()
  
  ls_notin_BTS <- ls %>% 
    filter(DB_sp %nin% ls_in_BTS) %>%
    mutate(inBTSdata = "not")
}
```


Running the functions  
```{r function_run_combine}
gen_ <- get_sp_in_taxon("Genus", Genus, DB_sp)
fam_ <- get_sp_in_taxon("Family", Family, DB_sp)
ord_ <- get_sp_in_taxon("Order", Order, DB_sp)
cls_ <- get_sp_in_taxon("Class", Class, DB_sp)

# variable name (nosp_yeshab): "not id'ed to species, yes species reps with habitat information"
nosp_yeshab <- 
  bind_rows(gen_, fam_, ord_, cls_) %>% 
  select(-BTS_sp) %>%
  mutate(BTS_sp = case_when(level == "Genus" ~ paste0(Genus, " sp"),
                            level == "Family" ~ Family,
                            level == "Order" ~ Order,
                            level == "Class" ~ Class,
                            TRUE ~ "NA") )

gen_no <- get_sp_notin_taxon("Genus", Genus)
fam_no <- get_sp_notin_taxon("Family", Family)
ord_no <- get_sp_notin_taxon("Order", Order)
cls_no <- get_sp_notin_taxon("Class", Class)

# variable name (nosp_nohab): "not id'ed to species, no species reps with habitat information"
nosp_nohab <- 
  bind_rows(gen_no, fam_no, ord_no, cls_no)

# removing detritus  
rm(gen_, fam_, ord_, cls_, gen_no, fam_no, ord_no, cls_no)
```

Taxa observations that do not have a species in the BTS data to represent them.  
```{r tb_not_sp_miss}
nosp_nohab %>%
  select(BTS_sp, Com_name, level, type) %>% 
  kable(col.names = c("BTS Observation", "Common name", "Level", "Type"),
        caption = "BTS observations that do not have species representatives in the BTS data") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"))
```

<br/>  


```{r}
# adding whether the non-species have a representative species or not 
select_list <- select_list %>% 
  mutate(sp_represent = ifelse(BTS_misspl %in% nosp_nohab$BTS_misspl, "no", "yes"))

# filtering the BTS data for all observations that are not species and do not have
# a species representative (aka, no habitat information to work with)
BTS_dta_nosp_nohab <- dta_ed %>% 
  filter(taxa %in% nosp_nohab$BTS_misspl)

# filtering the BTS data for all observations that are not species but do have a 
# species representative
BTS_dta_nosp_yeshab <- dta_ed %>% 
  filter(taxa %in% nosp_yeshab$BTS_misspl)

```

These are the counts of BTS observations that do not have a species representative.

```{r}
# nosp_no_dist_levels_1 <- c("Artediellus sp", "Penaeus sp", "Sphyraena sp",
#                          "Blenniidae", "Galatheidae",	"Gobiidae", "Majidae",
#                          "Myctophidae", "Pomacentridae", "Scaridae", "Stomatopoda")

nosp_no_dist_levels_2 <- c("Stomatopoda", "Scaridae", "Pomacentridae", "Myctophidae",
                           "Majidae", "Gobiidae", "Galatheidae", "Blenniidae",
                           "Artediellus sp", "Penaeus sp", "Sphyraena sp")

BTS_dta_nosp_nohab %>% 
  filter(!is.na(year)) %>%
  group_by(taxa, year) %>% 
  summarize(nb = sum(count)) %>% 
  ggplot(aes(fct_relevel(taxa, nosp_no_dist_levels_2), nb)) +
  geom_col() +
  facet_wrap(~ year) +
  coord_flip() +
  theme_bw() +
  labs(x = NULL,
       y = "Count",
       title = "BTS counts of organisms not classified to species",
       subtitle = "No species representative for the taxa in the data sets")

# This is a line graph of counts by year colored by taxa, but it's hard to read

# BTS_dta_nosp_nohab %>% 
#   filter(!is.na(year)) %>%
#   mutate(year = as.factor(year)) %>% 
#   group_by(taxa, year) %>% 
#   summarize(nb = sum(count)) %>%
#   ggplot(aes(year, nb, group = taxa, color = taxa)) +
#   geom_line() +
#   theme_bw()

```
<br/>  

I am a little surprised that there are no species representatives of the Family Myctophidae in the trawl data. I also think it is interesting that Myctophidae trawl counts jump from 324 in 2012 to 1176 in 2013 and stay above 500 through 2019. I wonder if that was part of the shelf warming that started in 2011 or if sampling shifted to the shelf break. 

I did not have time to make density surfaces of sampling effort and catches to check if sampling effort shifted or if myctophid distribution shifted, but I was able to plot out the myctophid catches in relation to all the sampled strata. The graph below is a heatmap of the amount of stations sampled in a stratum each year and the number of myctophids caught in each stratum.

It *looks* like there was a similar amount of sampling effort between 2012 and 2013 and there was an increase in the amount of myctophids caught in stratum 1080 in 2013. Looking at full time line, it looks like 2012 was the anomalous year in that no myctophids were caught in stratum 1080 but were caught in similar quantities in all other years. Again, I'm just eyeballing the graph and I have not plotted out which stations in each strata were sampled.
```{r}
BTS_dta_nosp_nohab %>% 
  filter(!is.na(year)) %>%
  group_by(taxa, year) %>% 
  summarize(nb = sum(count)) %>% 
  filter(taxa == "Myctophidae") %>% 
  kable(col.names = c("Taxa", "Year", "Total count")) %>% 
  kable_styling(bootstrap_options = c("striped", "hover"))


# BTS_dta_nosp_nohab %>% 
#   filter(taxa == "Myctophidae") %>%
#   filter(!is.na(year)) %>%
#   group_by(year, stratum) %>% 
#   summarize(nb = sum(count)) %>% 
#   ggplot(aes(factor(year), nb)) +
#   geom_col() +
#   facet_wrap(~ stratum, scales = "free_y") +
#   theme_bw() +
#   theme(axis.text.x = element_text(angle = 90, vjust = 0.25, hjust = 1))
```


```{r fig.height= 10, fig.width=10}

mycto <- BTS_dta_nosp_nohab %>% 
  filter(taxa == "Myctophidae") %>%
  group_by(year, stratum) %>% 
  summarize(nb = sum(count))


dta %>% 
  mutate(year = year(begindatetime)) %>% 
  group_by(year, stratum) %>% 
  summarize(nb_station = n()) %>% 
  ggplot() +
  geom_tile(aes(factor(year), factor(stratum), fill = nb_station)) +
  scale_fill_viridis_c(option = "magma") +
  geom_point(data = mycto, aes(factor(year), factor(stratum), size = nb), 
             color = "grey", alpha = 0.7) +
  theme_bw() +
  labs(x = "Year", 
       y = "Stratum",
       fill = "Number of\nstations",
       size = "Number of\n Myctophids",
       caption = "NA are the stations that do not have a datetime, longitude, or latititude"  )
```

```{r}
dta_ed %>% 
  filter(taxa == "Myctophidae",
         stratum == 1080) %>% 
  group_by(year) %>% 
  summarize(nb_mycto = sum(count)) %>%
  # Have to add a 0 for 2012 or it does not show up in the plot
  bind_rows(tibble(year = 2012, nb_mycto = 0)) %>% 
  ungroup() %>% 
  ggplot(aes(factor(year), nb_mycto)) +
  geom_col() +
  theme_bw() +
  labs(x = "Year",
       y = "Number",
       caption = "Number of myctophids caught in stratum 1080.")
  
```



## Habitat distributions of the species representatives

This is a two part process. First, we need to see what the habitat distribution of the species representatives for the upper level taxa look like. Second, we assign the upper level taxa habitat(s) based on the weights from the species representatives. 


I have not mapped out the data to see if the species representatives are found in the same strata as the upper level taxa they are associated with (do not forget that some strata listed in the BTS data DO NOT match the strata in the shapefile.).  
```{r fig.width = 8, fig.height = 7}

nosp_yeshab %>%
  count(BTS_sp, level, type) %>% 
  mutate(level = factor(level, levels = c("Genus", "Family", "Order", "Class"))) %>% 
  arrange(level) %>% 
  kable(col.names =  c("BTS Observation", "Level", "Type", 
                       "Number of species <br/>representatives"),
        caption = "Number of species representatives for BTS observations",
        escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover"))


BTS_dta_nosp_yeshab_taxa <- BTS_dta_nosp_yeshab %>% 
  left_join(nosp_yeshab, by = c("taxa"  = "BTS_misspl")) %>% 
  select(-LarvalArea, -StockCode, -Kingdom) %>% 
  mutate(level = factor(level, levels = c("Genus", "Family", "Order", "Class")))

habitat_percent <- BTS_dta_nosp_yeshab_taxa %>% 
  group_by(level, BTS_sp, habitat) %>% 
  summarize(nb_individuals = sum(count, na.rm = TRUE)) %>% 
  mutate(per = nb_individuals / sum(nb_individuals),
         id = cur_group_id())

plot_hab_per <- function(dta, id_start, id_stop){
  dta %>% 
  filter(id >= id_start & id <=id_stop) %>% 
  ggplot(aes(fct_reorder(BTS_sp, id), per, fill = habitat)) +
  geom_col(position = "fill") +
  geom_text(aes(label = nb_individuals), 
            position = "stack", vjust = 2.0, size = 3) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#3F4788FF", "#287D8EFF", "#3CBC75FF","#DCE318FF")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.25, hjust = 1),
        legend.position = "bottom") +
  labs(x = NULL,
       y = "Percent of species representatives",
       fill = NULL,
       caption = "Habitat distributions of the species representatives for the upper level taxa. \nNumbers are total trawl counts of species representatives. \nDoes not include the counts of the taxa.")
}

plot_hab_per(habitat_percent, 1, 19)
plot_hab_per(habitat_percent, 20, 39)
plot_hab_per(habitat_percent, 40, 57)

```

How you would like to assign the habitat percentages to the BTS counts of organisms not classified to species is up to you. :)

# Habitat for species counts

`species_cnts` is the data frame of habitat information added to the trawl data for all organisms classified to species. Below is a print out of all the column names and the first few observations.  

`taxa` is the name given in the BTS data. This column has misspellings in it.  
`DemersPelag` through `Kingdom` is information assigned by FishBase and SeaLifeBase.  
`GasBladdreEarly` and `GasBladderLate` can effectively be ignored as they only have to do with larval development.  
`habitat` is the paired-down version of FishBases `DemersPelag` (I assigned it).  


```{r}
species_cnts <- dta_ed %>% 
  filter(taxa %nin% select_list$BTS_misspl,
         !str_detect(taxa, "^Unknown")) %>% 
  left_join(select(both_fish_sealife, -SpecCode, -FamCode, -StockCode, -LarvalArea),
            by = c("taxa" = "BTS_sp"))

glimpse(species_cnts)
```

<br/>  


```{r}
species_cnts %>% 
  filter(!is.na(begindatetime)) %>% 
  group_by(habitat, cruiseid, type) %>%
  summarize(nb = sum(count)) %>% 
  ggplot(aes(habitat, nb, fill = type)) +
  geom_col(position = position_dodge2()) +
  scale_fill_viridis_d(labels = c("Fish", "Not Fish")) +
  scale_y_log10(labels = scales::comma) +
  facet_wrap(~ cruiseid) +
  coord_flip() +
  theme_bw() +
  theme(legend.position = "bottom") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL,
       y = "Count",
       fill = NULL,
       caption = "X-axis is on a log10 scale.")
```
<br/>

That bar chart is kind of annoying. Here's line graph.
```{r fig.width=8}
smry <- species_cnts %>% 
  group_by(cruiseid, habitat, type) %>%
  summarize(nb = sum(count)) %>% 
  ungroup() %>% 
  mutate(year = substring(cruiseid, 3, 6))

labels <- c(fish = "Fish", not_fish = "Not Fish")

smry %>% 
  ggplot(aes(year, nb, color = habitat, group = habitat)) +
  geom_line() +
  scale_color_manual(values = c("#3F4788FF", "#287D8EFF", "#3CBC75FF","#DCE318FF")) +
  scale_y_log10(labels = scales::comma) +
  facet_grid(~ type, labeller=labeller(type = labels)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL,
       y = "Count",
       color = "Habitat",
       caption = "Y-axis is on a log10 scale.")

  
```

<br/> 
