---
title: "EDA 1 - Cleaning species names via rfishbase"
author: "Erin LaBrecque"
date: "10/09/2020"
output: github_document
always_allow_html: true
---

This is Exploratory Data Analysis (EDA) file 1 of the NEFSC Bottom Trawl Survey catch counts from 2008 through 2019.  

I exposed all the R code in case you wanted to explore it. If you want to explore and run the code in these HTML files, I can send you the rMarkdown files (.Rmd) or you can copy and paste the code from here. You will need to create the folders, `output_input` and `output_csv` in your working directory to store the intermediate .rds and .csv files. You will use `output_input` in this EDA and both `output_input` and `output_csv` will be used in subsequent EDAs.

```{r setup, results='hide', message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      fig.width = 7, fig.height = 6.5)
library(knitr)
library(tidyverse)
library(fs)
library(lubridate)
library(kableExtra)
library(rfishbase)
```

# Loading the data

To successfully run this code, all the bottom trawl survey .csv files need to be in a folder called `data` that needs to be in your working directory. 

```{r readin, results='hide'}
# These lines of code tell R to look in the "data" folder, read in each .csv file,
# and then bind all the row from each file together. All columns are kept.
dta <- dir_ls(path = "data") %>% 
  lapply(read_csv, col_types=cols(begindatetime=col_datetime("%d/%m/%Y-%H:%M:%S"))) %>% 
  bind_rows()

# filtering rows with missing metadata
bad_dta <- dta %>% 
  filter(is.na(begindatetime))

# creating a metadata list to use later
meta_data <- c("cruiseid", "cruise", "station", "beginlon", "beginlat", "begindatetime", "stratum")
```

When reading in the data you will get warnings about parsing failures. This is because the `col_types` arguments above is looking for a **"%d/%m/%Y-%H:%M:%S"** datetime format for `begindatetime`, but there are several rows in a few of the .csv files that do not have data in the `begindatetime` column.

<br/>  
Example warning

:Warning: 2 parsing failures.

|row|col|expected|actual|file|
|---|-------------|---------------------------|---|-------------------------------------|
|227|begindatetime|date like %d/%m/%Y-%H:%M:%S| - |'data/HB201301_FSCS_Catch_output.csv'|
|373|begindatetime|date like %d/%m/%Y-%H:%M:%S| - |'data/HB201301_FSCS_Catch_output.csv'|  
<br/>

## Missing data from parse warnings

These `r nrow(bad_dta)` Cruise ID/Station/Stratum combinations do not have a `begindatetime`, `latitude`, or `longitude`. Species were collected at these stations. 

```{r missing_DT}
bad_dta %>% 
  select(cruiseid, cruise, station, stratum) %>% 
  kable(col.names = c("Cruise ID", "Cruise", "Station", "Stratum"),
               escape = FALSE) %>% 
   kable_styling(bootstrap_options = c("striped"), full_width = TRUE) 

# removing variable 
rm(bad_dta)
```
<br/>  

## Reorganizing the data frame

```{r pivot}
# making a list of taxa names from the column names
taxa_ls <- names(dta[8:ncol(dta)]) %>% 
  sort()

# pivoting the table from wide to long,
# putting taxa values in sentence case, adding year and month variables
dta_long <- dta %>% 
  pivot_longer(all_of(taxa_ls),
               names_to = "taxa", 
               values_to = "count",
               values_drop_na = TRUE) %>% 
  mutate(taxa  = str_to_sentence(taxa),
         year =  year(begindatetime),
         month = month(begindatetime, abbr = FALSE, label = TRUE))

# removing 0 counts to work with a smaller data set
dta_ed <- dta_long %>% 
  filter(count > 0)

# dta_long is almost 70 MB. Removing it.
rm(dta_long)
```
<br/>  

# Query FishBase and SeaLifeBase

Using the r package, `rfishbase`, we can access information in [FishBase](www.fishbase.org) without having to go through each species one-by-one. `rfishbase` also has an experimental feature to query [SeaLifeBase](www.sealifebase.ca). In the next few sections I query both databases using `species()` to determine if a given species and associated life history data can be retrieved programmatically.

<br/>  

First, separate the data into the different taxonomic groups. Life history information can only be retrieved for species.  
```{r taxa_separation}
# converting taxa list to sentence case
taxa_sentence_case <- str_to_sentence(taxa_ls)

# parsing out the higher taxonomy levels
poda <-  str_subset(taxa_sentence_case, pattern = "poda$")

formes <- str_subset(taxa_sentence_case, pattern = "formes$") 

fam <- str_subset(taxa_sentence_case, pattern = "idae$") 

unknws <- str_subset(taxa_sentence_case, pattern = "^Unknown")

genus <- str_subset(taxa_sentence_case, pattern = " sp$")

ptrn <- c("Trash species in catch", "Crustacea shrimp", 
          "Selachimorpha", "Stelleroidea")
outliers <- map(ptrn, str_subset, string = taxa_sentence_case) %>% 
  unlist()
  
species <- taxa_sentence_case[!taxa_sentence_case %in% 
                                c(poda, formes, fam, genus, unknws, outliers)]
```

- unknws = `r length(unknws)` 
- outliers = `r length(outliers)`  
- poda = `r length(poda)`  
- formes = `r length(formes)`  
- fam = `r length(fam)`  
- genus = `r length(genus)`  
- species = `r length(species)`  

<br/>  

```{r table_others}
kable(list(outliers, poda, formes, unknws), 
      col.names = NULL,
      caption = "Unknown and taxa above Family")  %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) 
```

<br/>   

```{r table_family}
fam %>% 
  as_tibble() %>% 
  split(1:3) %>% 
  kable(col.names = "Family",
        caption = "Families") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", full_width = TRUE)) 
```

<br/>    

```{r table_genus}
genus %>% 
  as_tibble() %>% 
  split(1:2) %>%  
  kable(col.names = "Genus",
        caption = "Genera") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", full_width = TRUE)) 
```

<br/>  

Now query the databases for species. This will take a minute or two if the databases have not been cashed.
```{r query}
# querying fishbase to see which taxa are fish
sp_fish <- species(species, fields = c("SpecCode", "Species", "FBname"), 
                   server = "fishbase") %>% 
  rename(BTS_sp = Species, Com_name = FBname)

# querying sealifebase to see which taxa are something other than fish
sp_sealife <- species(species, fields = c("SpecCode", "Species", "FBname"), 
                      server = "sealifebase") %>% 
  rename(BTS_sp = Species, Com_name = FBname)
```

<br/>  

This section checks that all species listed have been identified in one of the two databases.  
```{r combine_query_results}
# turning the species list into a data frame
sp_full <- tibble(species) %>% 
  rename(BTS_sp = species)

# combining the query results with the original list of species and renaming columns
# FB = fishbase, SLB = sealifebase
sp_checks <- bind_cols(sp_full, sp_fish, sp_sealife) %>% 
  select(-starts_with("Com_name")) %>% 
  rename(BTS_sp = BTS_sp...1,
         FB_SpecCode = SpecCode...2,
         FB_species = BTS_sp...3,
         SLB_SpecCode = SpecCode...5,
         SLB_species = BTS_sp...6)

# determining what entries in the species list were not in either database
not_in_either_db <- sp_checks %>% 
  filter(is.na(FB_species) & is.na(SLB_species)) %>% 
  select(BTS_sp)
```
<br/>  

`r nrow(not_in_either_db)` species in the bottom trawl data set were not found in either database using the function `species()`.
```{r table_species}
not_in_either_db %>% 
  as_tibble() %>% 
  split(1:3) %>%
  kable(caption = "Species in the BTS data not found in either database") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", full_width = TRUE)) %>%
  column_spec(1, italic = TRUE) %>% 
  column_spec(2, italic = TRUE) %>% 
  column_spec(3, italic = TRUE)
```
<br/>  

### Checking for synonoms  
The species listed above might be in the databases under different names. Thankfully `rfishbase` has a function that looks up synonyms.  

```{r synonyms}
# checking for synonyms in fishbase, removing NAs, removing any duplicates
sp_fish_sy <- synonyms(pull(not_in_either_db), server = "fishbase") %>% 
  select(BTS_sp = synonym, Status, SpecCode, TSN, DB_sp = Species) %>% 
  filter(!is.na(TSN)) %>% 
  distinct(DB_sp, .keep_all = TRUE)

# checking for synonyms in sealifebase and removing NAs, removing any duplicates
sp_sealife_sy <- synonyms(pull(not_in_either_db), server = "sealifebase") %>% 
  select(BTS_sp = synonym, Status, SpecCode, DB_sp = Species) %>% 
  filter(!is.na(Status)) %>% 
  distinct(DB_sp, .keep_all = TRUE)

# checking to see if any species were missed
if (nrow(sp_fish_sy) + nrow(sp_sealife_sy) == nrow(not_in_either_db)){
  print("All species acounted for!")
} else {
  paste(nrow(not_in_either_db) - (nrow(sp_fish_sy) + nrow(sp_sealife_sy)), "species not accounted for.")
}
```

<br/>

### Finding the species that did not have synonyms

```{r rogue_species}
no_syn_sp <- bind_rows(sp_fish_sy, sp_sealife_sy) %>% 
  select(BTS_sp)

not_in_either_db %>% 
  filter(!BTS_sp %in% no_syn_sp$BTS_sp) %>% 
  kable(col.name = "BTS name",
        caption = "Species in the BTS data not found as synonyms the databases") %>% 
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  column_spec(1, italic = TRUE)
```
<br/>  

Looks like a of couple misspellings (or a character limit was set in the bottom trawl survey database), not synonym issues. This section fixes the misspelling but keeps the original spelling in `BTS_sp` column to reference back to the original data. 
```{r spelling_fix}
sp_spell_fix <- species(c("Urophycis earllii", "Aluterus heudelotii"), 
                  fields = c("SpecCode", "Species", "FBname"), server = "fishbase") %>% 
  rename(DB_sp = Species, Com_name = FBname) %>% 
  mutate(BTS_sp = case_when(DB_sp == "Urophycis earllii" ~ "Urophycis earlli",
                            DB_sp == "Aluterus heudelotii" ~ "Aluterus heudeloti"))


#combining data frames
sp_syn <- bind_rows(sp_fish_sy, sp_sealife_sy)

# printing out table
sp_syn %>%
  select(BTS_sp, Status, DB_sp) %>%
  kable(caption = "Synonyms found in FishBase and SeaLifeBase",
  col.names = c("BTS name", "Status", "Database name")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", full_width = TRUE)) %>%
  column_spec(1, italic = TRUE) %>%
  column_spec(3, italic = TRUE)
```

<br/>   
 
This section queries FishBase and SeaLifeBase to get the species codes and common names for the synonyms.  
```{r query_2}
# getting common names and species codes for FishBase then joining to the synonym data frame
# to attach the original names
sp_fish_sy_spcodes <- species(sp_fish_sy$DB_sp,  
                              fields = c("SpecCode", "Species", "FBname"), 
                              server = "fishbase") %>% 
  rename(DB_sp = Species, Com_name = FBname) %>% 
  left_join(sp_fish_sy, by = "SpecCode")%>% 
  select(SpecCode, BTS_sp, Status, Com_name, DB_sp = DB_sp.x, TSN)

# same for sealife base
sp_sealife_sy_spcodes <- species(sp_sealife_sy$DB_sp,  
                              fields = c("SpecCode", "Species", "FBname"), 
                              server = "sealifebase") %>% 
  rename(DB_sp = Species, Com_name = FBname) %>% 
  left_join(sp_sealife_sy, by = "SpecCode")%>% 
  select(SpecCode, BTS_sp, Status, Com_name, DB_sp = DB_sp.x)
```

### Combing data frames

```{r}
# removing the NAs (not fish) from the original fish query and adding in the new spellings
sp_fish_almost_all <- sp_fish %>% 
  filter(!is.na(BTS_sp)) %>% 
  bind_rows(sp_spell_fix) %>% 
  mutate(DB_sp = ifelse(is.na(DB_sp), BTS_sp, DB_sp))

sp_fish_all <- sp_fish_almost_all %>% 
  bind_rows(sp_fish_sy_spcodes)

# same for sealifebase but do not need to add in the misspellings
sp_sealife_all <- sp_sealife %>% 
  filter(!is.na(BTS_sp)) %>% 
  mutate(DB_sp = BTS_sp) %>% 
  bind_rows(sp_sealife_sy_spcodes)
```

Species names have been checked, but the common names have not. This section checks the common names for all the species.

```{r com_name_check}
sp_sealife_all %>% 
  filter(is.na(Com_name)) %>% 
  select(-SpecCode) %>% 
  kable(col.names = c("BTS name", "Common name", "Database name", "Status"),
        caption = "Missing common name") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", full_width = TRUE)) %>% 
  column_spec(c(1,3), italic = TRUE)


sp_fish_all %>% 
  filter(is.na(Com_name)) %>% 
  select(-SpecCode, -TSN) %>% 
  kable(col.names = c("BTS name", "Common name", "Database name", "Status"),
        caption = "Missing common name") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", full_width = TRUE)) %>% 
  column_spec(c(1,3), italic = TRUE)
```
<br/>

Adding the common names to the data.

```{r fix_common_name}
sp_sealife_all <- sp_sealife_all %>% 
  mutate(Com_name = case_when(BTS_sp == "Pandalus propinquus" ~ "Shrimp",
                              TRUE ~ Com_name))

sp_fish_all <- sp_fish_all %>% 
  mutate(Com_name = case_when(BTS_sp == "Polymetme thaeocoryla" ~ "Lightfishes",
                              TRUE ~ Com_name))
```


<br/>  

# Wrap up EDA 1

This file is getting long. I'll save the outputs here an start a new file to look up habitats and gas bladder information. Un-comment (remove #) to write out the data files.
```{r save}
# write_rds(dta, "output_input/dta.rds")
# write_rds(dta_ed, "output_input/dta_ed.rds")
# write_rds(sp_sealife_all, "output_input/sp_sealife.rds")
# sp_fish_all %>% 
#   select(-TSN) %>% 
#   write_rds("output_input/sp_fish.rds")

```

